Mastery Learning in Practice:\\ A Descriptive Analysis of Data from the Cognitive Tutor Algebra I Effectiveness Trial

### 2017-10-04 paper.Rnw
progress....

\section{Standard versus Customized Curricula}

All students who participated in the treatment condition of the study were enrolled in at least one curriculum throughout the school year. In Year 1, the standard Algebra curriculum was the most-used curriculum across schools participating in the study; other curricula were Bridge-to-Algebra (pre-algebra), Algebra II, and five customized curricula that were used in two schools. The standard curricula used in Year 2 included the three that were used in Year 1, plus Geometry and Algebra II Bonus. In Year 2, Carnegie Learning also developed many more customized curricula with input from participating schools, for a total of 33 curricula used in 13 schools. This led to fewer students enrolled in the standard curriculum and more enrolled in customized curricula overall in Year 2.

In the available data, representing a subset of the entire study sample, the number of students enrolled in at least one standard curriculum in Year 1 was 2,801; in Year 2, this number decreased to 1,204. The number of students enrolled in at least one customized curriculum was 115 in Year 1 and increased to 2,333 in Year 2. It should be noted that students could be enrolled in multiple curricula. For instance, teachers could choose to enroll students in any of the standard curricula provided, such as Algebra I or Algebra II, along with any of the customized curricula that were used by the school.

The use of customized curricula varied substantially by state. For instance, schools in CT and NJ did not use customized curricula in either Year 1 or Year 2. All other states contained schools that used customized curricula in addition to the standard curricula. CT schools maintained similar number of students in Algebra I across both years with 111 in Year 1 and 135 in Year 2. NJ substantially decreased the number of students in the CTAI program from 89 to 32 from Year 1 to Year 2. However, while all other states (KY, LA, MI, and TX) did decrease enrollment in the standard Algebra I curriculum the second year, those same states increased enrollment in the customized curricula. KY added 3 customized curricula with 146 students, LA added 15 customized curricula with 286 students, MI added 5 customized curricula with 370 students, and TX added 10 additional customized curricula and had the largest enrollment in these curricula with 1,531 students. Students enrolled in the standard Algebra I curriculum dropped from 974 in Year 1 to 61 in Year 2 in TX. As noted above, students could be enrolled in more than one curriculum. In Year 2, the average number of curricula per a student ranged from 1.00 (NJ and CT) to 2.53 (TX). The next highest value was LA at 2.10.

@@

\section{Student Usage Across Sections and Years}
\begin{figure}
<<usageTime, fig.height=3,fig.width=6>>=
secByStud <- summarize(group_by(data,state,Yr,field_id),numSec=n_distinct(section,na.rm=TRUE),time=sum(total_t1,na.rm=TRUE),mastered=n_distinct(section[status=='graduated'],na.rm=TRUE))

@@

print(timeStateYear <- ggplot(secByStud,aes(Yr,time/3600000))+geom_boxplot(outlier.shape=NA)+facet_grid(~state)+coord_cartesian(ylim=c(0,100))+labs(x='',y='Time on CTAI Software per Student (Hours)'))
#ggsave('timeStatYear.jpg',width=6,height=3)

@
\caption{Boxplots of hours each student spent on Cognitive Tutor
  software over by year and state. Students spending more than 100
  hours (\Sexpr{sum(secByStud[['time']]>100,na.rm=TRUE)} of \Sexpr{nrow(secByStud)}) and students
  with no timestamp data (\Sexpr{sum(is.na(secByStud[['time']]))})
  were excluded.}
\label{fig:timeByStud}
\end{figure}

\begin{figure}
<<mastSec,dependson='usageTime'>>=
print(mastStateYear <- ggplot(secByStud,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
@
\caption{Boxplots of the number of sections each of Cognitive Tutor
  software each student mastered, by year and state. Students
  mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)}) of \Sexpr{nrow(secByStud)}) and students
  with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))})
  were excluded.}
\label{fig:mastSec}
\end{figure}

Figure \ref{fig:timeByStud} shows that the number of sections worked varied substantially between students and across states and years. Across years, students in Texas, New Jersey, and Connecticut worked fewer sections each than in Kentucky, Louisiana, and Michigan. From year 1 to 2, the median number of sections a student worked on decreased in every state, decreasing more substantially Texas and New Jersey than in others. The variance of the numbers of sections students worked in Michigan increased from year 1 to 2 as well.

Figure 2 shows the proportion of worked sections that were from a customized curriculum in the three states that used customized curricula substantially: Kentucky, Michigan, and Texas. As students worked fewer sections in year 2, a greater proportion of the sections they did work were from customized curricula. Across states, the average number of sections worked per student enrolled for the standard curricula was 53.83 in Year 1 and was similar at 52.45 in Year 2 across all states. For customized curricula in Year 2, the average number of worked sections per student enrolled was 19.93. However, since students were often enrolled in multiple curricula, these numbers underestimate the number of sections worked by students over the whole year.

@@

>>>>

\section{Standard and Customized Curricula}

\begin{figure}
  \centering
<<curricula,dependson='data',fig.width=6.4,fig.height=3>>=

omar <- par()$oma

par(oma=c(0,2,0,0),mar=c( 3,.5,2,0))
par(mfrow=c(1,9))
for(st in levels(data$State)){
#    cwy <- currWorkedSt%>%filter(State==st)%>%group_by(Curriculum,overall)%>%summarize(`Yr 1`=p[Year=='Year 1'],`Yr 2`=p[Year=='Year 2'])
    cwy <- data%>%filter(!is.na(Curriculum) & State==st)%>%group_by(Curriculum,overall)%>%
    summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
    for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
    cwy <- cwy[nrow(cwy):1,]
    barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',75,NA),main=st,yaxt='n')
    if(st==levels(data$State)[1]) axis(2,at=seq(0,1,0.1),labels=paste0(seq(0,100,10),'\\%'))
    #box()
}
cwy <- data%>%filter(!is.na(Curriculum))%>%group_by(Curriculum,overall)%>%
    summarize(`Yr 1`=sum(Year=='Year 1'),`Yr 2`=sum(Year=='Year 2'))
for(y in c(1,2)) cwy[,paste('Yr',y)] <- cwy[,paste('Yr',y)]/sum(cwy[,paste('Yr',y)])
cwy <- cwy[nrow(cwy):1,]
barplot(as.matrix(cwy[,3:4]),col=COLS[as.numeric(cwy$Curriculum)],angle=ifelse(cwy$overall=='Customized',45,NA),density=ifelse(cwy$overall=='Customized',75,NA),main='Overall',yaxt='n')

plot(1:10,type='n',xaxt='n',yaxt='n',bty='n')
legend(-1.5,7.5, c("Bridge-to-\n Algebra",'Algebra I','$>$Alg. I','','Standard','Cust.'),
       fill=c(COLS,NA,'#010101','#010101'),border=rep('white',6),
       angle=c(NA,NA,NA,NA,NA,45),density=c(NA,NA,NA,NA,NA,75),xpd = TRUE, bty = "n",inset=c(0,0))

par(oma=omar)
@
\caption{Percentage of worked problems coming from various levels
  (denoted by color, with Algebra II and Geometry bundled as ``$>$Algebra I''), from
  standard and customized variants, denoted by shading.}
\label{fig:curricula}
\end{figure}

Students' automatic progress throught the CTAI software is governed by
curricula---sequences of sections and units embedded in the software.
Without external meddling, the curriculum a student works on
determines what section he or she will be directed towards next after
mastering (or exhausting the problems) from a previous section.
In the CTAI effectiveness trial, the most common curriculum was,
naturally, Algebra I.
This came with three closely related varients, due to software releases.
Students requiring more remediation were able to work on a less
advanced curriculum, called ``Bridge to Algebra,'' and more advanced
students could work on Algebra II or Geometry.

In the second year of the study, some high schools, primarily in
Texas, Michigan, and Kentucky, requested customized variants of the
curricula.
This was typically due to state standards and testing schedules.
These ``customized curricula'' altered the order of some sections and
units, and were particular to schools.

Figure \ref{fig:curricula} shows the percentage of worked problems
from each curriculum, from standard and customized varieties, by state
and year.
First, note that the vast majority of worked problems were from the
Algebra I sequence.
A small but notable number of more advanced problems were worked in
Kentucky in Year 2, and some less advanced problems were worked in
Michigan and Louisiana.
Secondly, note the rise in ``customized curricula'' in year 2 in
Texas, Kentucky, and Michigan, the three states with the most students
in our dataset.
In particular, Texas shifted almost entirely to customized curriclua
from years 1 to 2.

% All students who participated in the treatment condition of the study were enrolled in at least one curriculum throughout the school year.
% Figure
% In Year 1, the standard Algebra curriculum was the most-used curriculum across schools participating in the study; other curricula were Bridge-to-Algebra (pre-algebra), Algebra II, and five customized curricula that were used in two schools. The standard curricula used in Year 2 included the three that were used in Year 1, plus Geometry and Algebra II Bonus. In Year 2, Carnegie Learning also developed many more customized curricula with input from participating schools, for a total of 33 curricula used in 13 schools. This led to fewer students enrolled in the standard curriculum and more enrolled in customized curricula overall in Year 2.

% In the available data, representing a subset of the entire study sample, the number of students enrolled in at least one standard curriculum in Year 1 was 2,801; in Year 2, this number decreased to 1,204. The number of students enrolled in at least one customized curriculum was 115 in Year 1 and increased to 2,333 in Year 2. It should be noted that students could be enrolled in multiple curricula. For instance, teachers could choose to enroll students in any of the standard curricula provided, such as Algebra I or Algebra II, along with any of the customized curricula that were used by the school.

% The use of customized curricula varied substantially by state. For instance, schools in CT and NJ did not use customized curricula in either Year 1 or Year 2. All other states contained schools that used customized curricula in addition to the standard curricula. CT schools maintained similar number of students in Algebra I across both years with 111 in Year 1 and 135 in Year 2. NJ substantially decreased the number of students in the CTAI program from 89 to 32 from Year 1 to Year 2. However, while all other states (KY, LA, MI, and TX) did decrease enrollment in the standard Algebra I curriculum the second year, those same states increased enrollment in the customized curricula. KY added 3 customized curricula with 146 students, LA added 15 customized curricula with 286 students, MI added 5 customized curricula with 370 students, and TX added 10 additional customized curricula and had the largest enrollment in these curricula with 1,531 students. Students enrolled in the standard Algebra I curriculum dropped from 974 in Year 1 to 61 in Year 2 in TX. As noted above, students could be enrolled in more than one curriculum. In Year 2, the average number of curricula per a student ranged from 1.00 (NJ and CT) to 2.53 (TX). The next highest value was LA at 2.10.

@@

\section{Student Usage Across Sections and Years}
\begin{figure}
<<usageTime, fig.height=3,fig.width=6,dependson='data'>>=
secByStud <- summarize(group_by(data,state,Yr,field_id),numSec=n_distinct(section,na.rm=TRUE),time=sum(total_t1,na.rm=TRUE),mastered=n_distinct(section[status=='graduated'],na.rm=TRUE))

secByStud$time <- secByStud$time/3600000

@@

levels(secByStud$state) <- c(levels(secByStud$state),'Overall')
secByStud2 <- secByStud
secByStud2 <- rbind(secByStud,secByStud2)
secByStud2$state[seq(nrow(secByStud)+1,2*nrow(secByStud))] <- 'Overall'

print(timeStateYear <- ggplot(secByStud2,aes(Yr,time))+geom_boxplot(outlier.shape=NA)+facet_grid(~state)+coord_cartesian(ylim=c(0,110))+labs(x='',y='Hours on CTAI Software per Student'))
#ggsave('timeStatYear.jpg',width=6,height=3)

@
\caption{Boxplots (excluding outliers) of hours each student spent on Cognitive Tutor
  software over by year and state. Students  with no timestamp data
  (\Sexpr{sum(is.na(secByStud[['time']]))/2})
  were excluded.}
\label{fig:timeByStud}
\end{figure}

\begin{figure}
<<mastSec,dependson='usageTime',fig.width=6,fig.height=3,dependson=c('usageTime','data')>>=
print(mastStateYear <- ggplot(secByStud2,aes(Yr,mastered))+geom_boxplot()+facet_grid(~state)+labs(x='',y='\\# Sections Mastered Per Student')+coord_cartesian(ylim=c(0,250)))
@
\caption{Boxplots of the number of sections each of Cognitive Tutor
  software each student mastered, by year and state. Students
  mastering more than 250 sections (\Sexpr{sum(secByStud[['mastered']]>250,na.rm=TRUE)/2}) of \Sexpr{nrow(secByStud)/2}) and students
  with no mastery data (\Sexpr{sum(is.na(secByStud[['mastered']]))/2})
  were excluded.}
\label{fig:mastSec}
\end{figure}

Figure \ref{fig:timeByStud} shows that the number of hours students
spent working on the CTAI software
varied substantially between students and across states and
years.
Overall, students spent less time on the software in Year 2 than in
Year 2.
Specifically, in the first year the students spent a median of
\Sexpr{round(median(secByStud[['time']][secByStud[['Yr']]=='Yr 1'],na.rm=TRUE),1)} hours on the software, compared to \Sexpr{round(median(secByStud[['time']][secByStud[['Yr']]=='Yr 2'],na.rm=TRUE),1)} in  the second year.
However, usage varied more in year 2 as well---the interquartile
range of time spent was \Sexpr{round(IQR(secByStud[['time']][secByStud[['Yr']]=='Yr 1'],na.rm=TRUE),1)} hours in the first year, compared to \Sexpr{round(IQR(secByStud[['time']][secByStud[['Yr']]=='Yr 2'],na.rm=TRUE),1)} in  the second year.

Time spent varied substantially by state, with students in Texas,
Connecticut, and New Jersey working far fewer hours than students in
Kentucky, Louisiana, and Michigan.
Not every state reduced its usage from years 1 to 2---while students
in Texas, Kentucky and New Jersey used the software less in the second
year than in the first, students in Michigan, Louisiana, and
Connectuct increased their usage.

@@

\end{document}
